---
title: "Linear models"
output:
  html_document: 
    toc: true
    toc_float: true
---

Linear regression models are fundamental in statistics and data science. When seeking to understand how covariates are associated with outcomes, linear models are among the first, best options. Although other regression approaches are possible, the flexibility and  interpretability and of linear models make them essential. 

This is the first module in the [Linear Models](topic_linear_models.html) topic; the relevant slack channel is [here](https://p8105-fall2018.slack.com/messages/CCAAE9AAU).

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

library(tidyverse)

theme_set(theme_bw() + theme(legend.position = "bottom"))
```

## Some slides

<script async class="speakerdeck-embed" data-id="285289b17d194a4282d53f1800d37199" data-ratio="1.77777777777778" src="//speakerdeck.com/assets/embed.js"></script>
<div style="margin-bottom:5px"> <strong> <a href="https://speakerdeck.com/jeffgoldsmith/p8105-simulation" title="Simulation and Bootstrapping" target="_blank">Simulation</a> </strong> from <strong><a href="https://speakerdeck.com/jeffgoldsmith" target="_blank">Jeff Goldsmith</a></strong>. </div><br>


## Example

I'll write code for today's content in a new R Markdown document called `simulation.Rmd` in the `iteration` directory / repo. The code chunk below loads the usual packages and sets a seed for reproducibility.

```{r}
library(tidyverse)

set.seed(1)
```

### asdf


## Other materials

* 

The code that I produced working examples in lecture is [here](https://github.com/P8105/linear_models).
