---
title: "Simulation and Bootstrapping"
output:
  html_document: 
    toc: true
    toc_float: true
---

R's data structures, especially data frames, are surprisingly flexible. This is useful when the "observations" you want to store become more complex than single values; for example, each row many contain a few scalar observations as well a complete data set. In these cases, _list columns_ are an appropriate column type, and `map` functions provide a way to interact with those columns.

Bootstrapping is a popular resampling-based approach to statistical inference, and is helpful when usual statistical methods are intractable or inappropriate. The idea is to draw repeated samples from your original sample _with replacement_, thereby approximating the repeated sampling framework. Using list columns to store bootstrap samples is natural and provides a "tidy" approach to resampling-based inference.

This is the third module in the [Iteration](topic_iteration.html) topic; the relevant slack channel is [here](https://p8105-fall2018.slack.com/messages/CCAAE7XDW).

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)
```


## Some slides

<script async class="speakerdeck-embed" data-id="285289b17d194a4282d53f1800d37199" data-ratio="1.33333333333333" src="//speakerdeck.com/assets/embed.js"></script>
<div style="margin-bottom:5px"> <strong> <a href="https://speakerdeck.com/jeffgoldsmith/p8105-simulation-and-bootstrapping" title="Simulation and Bootstrapping" target="_blank">Simulation and Bootstrapping</a> </strong> from <strong><a href="https://speakerdeck.com/jeffgoldsmith" target="_blank">Jeff Goldsmith</a></strong>. </div><br>


## Example

I'll write code for today's content in a new R Markdown document called `listcols_and_bootstrapping.Rmd` in the `example_iteration` directory / repo. The code chunk below loads the usual packages.

```{r}
library(tidyverse)

theme_set(theme_bw())
theme_update(legend.position = "bottom")

set.seed(1)
```

Things are gonna get a little weird...

### Simulation: SLR for one $n$

Last class we wrote a short function to simulate data from a simple linear regression, fit the regression model, and return estimates of regression coefficients. Specifically, we generate data from 
$$ y_i = \beta_0 + \beta_1 x_i + \epsilon_i $$
for subjects $1 \leq i \leq n$ with $\epsilon_i \sim N[0,1]$ and return estimates $\hat{\beta}_0, \hat{\beta}_1$. That function is below. 

```{r sim_regression, cache = TRUE}
sim_regression = function(n, beta0 = 2, beta1 = 3) {
  
  sim_data = tibble(
    x = rnorm(n, mean = 1, sd = 1),
    y = beta0 + beta1 * x + rnorm(n, 0, 1)
  )
  
  ls_fit = lm(y ~ x, data = sim_data)
  
  tibble(
    beta0_hat = coef(ls_fit)[1],
    beta1_hat = coef(ls_fit)[2]
  )
}
```

Important statistical properties of estimates $\hat{\beta}_0, \hat{\beta}_1$ are established under the conceptual framework of repeated sampling. If you could draw from a population over and over, your estimates will have a known mean and variance:
$$ \hat{\beta}_0 \sim \left[\beta_0, \sigma^2 \left(\frac{1}{n} + \frac{\bar{x}}{\sum (x_i - \bar{x})^2}\right) \right] \mbox{ and } \hat{\beta}_1 \sim \left[\beta_1,\frac{\sigma^2}{\sum (x_i - \bar{x})^2} \right] $$
(Because our simulation design generates errors from a Normal distribution we also know that the estimates follow a Normal distribution, although that's not guaranteed by least squares estimation.)

In the real world, drawing samples is time consuming and costly, so "repeated sampling" remains conceptual. On a computer, though, drawing samples is pretty easy. That makes simulation an appealing way to examine the statistical properties of your estimators. 

Let's run `sim_regression()` 100 times to see the effect of randomness in $\epsilon$ on estimates $\hat{\beta}_0, \hat{\beta}_1$.

```{r loop_sim, cache = TRUE}
output = vector("list", 100)

for (i in 1:100) {
  output[[i]] = sim_regression(30)
}

sim_results = bind_rows(output)
```

Taking a look at the `for` loop we used to create these results, you might notice that there's no `input` list -- the sequence is used to keep track of the output but doesn't affect the computation performed inside the `for` loop. In cases like these, the `purrr::rerun` function is very handy. 

```{r}
sim_results = 
  rerun(100, sim_regression(30, 2, 3)) %>% 
  bind_rows()
```

Structurally, `rerun` is a lot like `map` -- the first argument defines the amount of iteration and the second argument is the function to use in each iteration step. As with `map`, we've replaced a for loop with a segment of code that makes our purpose much more transparent but both approaches give the same results. 

Let's make some quick plots and compute some summaries for our simulation results.

```{r}
sim_results %>% 
  ggplot(aes(x = beta0_hat, y = beta1_hat)) + 
  geom_point()

sim_results %>% 
  gather(key = parameter, value = estimate, beta0_hat:beta1_hat) %>% 
  group_by(parameter) %>% 
  summarize(emp_mean = mean(estimate),
            emp_var = var(estimate)) %>% 
  knitr::kable(digits = 3)
```

This is **_great_**! We've seen how our estimates are distributed under our simulation scenario, and can compare empirical results to theoretical ones. In this way, we can build intution for fundamental statistical procedures under repeated sampling in a way that's not possible with single data sets. 


### Simulation: SLR for several $n$s

Sample size makes a huge difference on the variance of estimates in SLR (and pretty much every statistical method). Let's try to clarify that effect through simulating at a few sample sizes. 

I'll start this process with a for loop around the code I established above using `rerun` (I could start from scratch by nesting one for loop in another for loop, but let's not). 

```{r}
n_list = list("n_30"  = 30, 
              "n_60"  = 60, 
              "n_120" = 120, 
              "n_240" = 240)
output = vector("list", length = 4)

for (i in 1:4) {
  output[[i]] = rerun(100, sim_regression(n_list[[i]])) %>% 
    bind_rows
}
```

After this loop, `output` is a list of 4 data frames; each data frame contains the results of 100 simulations at different sample sizes. 

Before we spend time looking at the results of the simulation, let's recast this using `map`. I want to use a single function in my call to `map_df`, so I'm going to write a wrapper for the call to `rerun` that allows me to change the parameters of the simulation (i.e. the argument to `sim_regression`) and the number of simulation replicates (i.e. the first argument to `rerun`). Once I have this, I'll call `map_df` to perform the complete simulation.

```{r, cache = TRUE, dependson=sim_regression}
simulate_n_regressions = function(n_runs = 100, n, beta0 = 2, beta1 = 3) {
  
  rerun(n_runs, sim_regression(n, beta0, beta1)) %>% 
    bind_rows()
  
}

sim_results = 
  map_df(.x = n_list, ~ simulate_n_regressions(n = .x), .id = "sample_size") 
```

Using a different call, I could increase the number of simulation runs or vary the parameters in the regression model:

```{r}
sim_results = 
  map_df(.x = n_list, ~ simulate_n_regressions(n = .x, n_runs = 1000, beta0 = 2, beta1 = 3),
         .id = "sample_size") 
```


Let's take a look at what we've accomplished in our simulations! First I'll take a look at the distribution of slope estimates across sample sizes. 

```{r}
sim_results %>% 
  mutate(sample_size = fct_inorder(sample_size)) %>% 
  ggplot(aes(x = sample_size, y = beta1_hat, fill = sample_size)) + 
  geom_violin()
```

These estimates are centered around the truth (3) for each sample size, and the width of the distribution shrinks as sample size grows. 

Next, I'll look at the bivariate distribution of intercept and slope estimates across sample sizes.

```{r}
sim_results %>% 
  mutate(sample_size = fct_inorder(sample_size)) %>% 
  ggplot(aes(x = beta0_hat, y = beta1_hat)) + 
  geom_point(alpha = .2) + 
  facet_grid(~sample_size)
```

The variability in the slope estimates was shown in the violin plot, but now we have a sense for the bivariate distribution of intercepts and slopes. Estimates of the intercept and slope are correlated with each other; this is expected from theoretical results describing the joint distribution of estimated regression coefficients. 

Lastly I'll look at the empirical mean and variance of these estimates.

```{r}
sim_results %>% 
  gather(key = parameter, value = estimate, beta0_hat:beta1_hat) %>% 
  group_by(parameter, sample_size) %>% 
  summarize(emp_mean = mean(estimate),
            emp_var = var(estimate)) %>% 
  knitr::kable(digits = 3)
```

These values are consistent with the formulas presented above. This kind of check is a useful way to support derivations (although they don't serve as a formal proof in any way).





### Bootstrapping

Bootstrapping is based on the idea of repeated sampling which underlies most approaches to statistical inference. Traditionally, the distribution of a sample statistic (sample mean, SLR coefficients, etc.) for repeated, random draws from a population has been established theoretically. These theoretical distributions make some assumptions about the underlying population from which samples are drawn, or depend on large sample sizes for asymptotic results.

In cases where the assumptions aren't met, or sample sizes aren't large enough for asymptotics to kick in, it is still necessary to make inferences using the sample statistic. In these cases, drawing repeatedly from the original population would be great -- one could simple draw a lot of samples and look at the empirical (rather than theoretical) distribution. But, as we said in [iteration and simulation](iteration_and_simulation.html), repeated sampling just doesn't happen in the real world.

Repeated sampling _can_ happen on a computer though. To bootstrap, one draws repeated samples (with the same sample size) from the original sample **_with replacement_** to mimic the process of drawing repeated samples from the population. The bootstrap samples will differ from the original sample, and the sample statistic of interest (sample mean, SLR coefficients, etc.) can be computed for each bootstrap sample. Looking at the distribution of the statistic across samples gives a sense of the uncertainty in the estimate. 

### Bootstrapping in SLR

Let's look at a couple of simulated data sets. Both are generated from a simple linear regression, but they have different error distributions. 

```{r}
set.seed(10)

n_samp = 250

sim_df_const = tibble(
  x = rnorm(n_samp, 1, 1),
  error = rnorm(n_samp, 0, 1),
  y = 2 + 3 * x + error
)

sim_df_nonconst = sim_df_const %>% 
  mutate(
  error = error * .75 * x,
  y = 2 + 3 * x + error
)

bind_rows(
  mutate(sim_df_const, data = "sim_df_const"),
  mutate(sim_df_nonconst, data = "sim_df_nonconst")
) %>% 
  ggplot(aes(x = x, y = y)) + 
  geom_point(alpha = .5) +
  stat_smooth(method = "lm") +
  facet_grid(~data) 
```

These datasets have roughly the same overall variance, but the left panel shows data with constant variance and the right panel shows data with non-constant variance. For this reason, ordinary least squares should provide reasonable estimates in both cases, but inference is standard inference approaches may only be justified for the data on the left.

The output below shows results from fitting simple linear regressions to both datasets.

```{r}
lm(y ~ x, data = sim_df_const) %>% summary()
lm(y ~ x, data = sim_df_nonconst) %>% summary()
```

Despite the very different error structures, standard errors for coefficient estimates are similar in both cases! 

We'll use the bootstrap to make inference for the data on the right. This is intended largely as an illustration for how to use the bootstrap in cases where the theoretical distribution is "unknown", although for these data in particular weighted least squares is more appropriate.


### Drawing one bootstrap sample

Let's write a quick function to generate our bootstrap samples. This function should have the data frame as the argument, and should return a sample from that dataframe drawn with replacement.

```{r}
boot_sample = function(df) {
  sample_frac(df, replace = TRUE)
}
```

We should also do a quick check to see if this is working.

```{r}
boot_sample(sim_df_nonconst)
```

That looks about right. 

### Drawing many bootstrap samples

We're going to draw repeated samples with replacement, and then analyze each of those samples separately. It would be really great to have a data structure that makes it possible to keep track of everything. Maybe a **_list column_**??!

Let's give that a try:

```{r}
boot_straps = data_frame(
  strap_number = 1:1000,
  strap_sample = rerun(1000, boot_sample(sim_df_nonconst))
)

boot_straps
```

We can do a few of quick checks to make sure this has worked as intended. First we'll look at a couple of bootstrap samples. 

```{r}
boot_straps %>% 
  filter(strap_number %in% 1:2) %>% 
  mutate(strap_sample = map(strap_sample, ~arrange(.x, x))) %>% 
  pull(strap_sample)
```

Seems okay -- some values are repeated, some don't appear in both datasets. Next I'll use ggplot to show some of these datasets, and to include a linear fit for each.

```{r}
boot_straps %>% 
  filter(strap_number %in% 1:3) %>% 
  unnest() %>% 
  ggplot(aes(x = x, y = y)) + 
  geom_point(alpha = .5) +
  stat_smooth(method = "lm", se = FALSE) +
  facet_grid(~strap_number) 
```

This shows some of the differences across bootstrap samples, and shows that the fitted regression lines aren't the same for every bootstrap sample.


### Analyzing bootstrap samples

My goal, of course, isn't to analyze bootstrap samples by plotting them -- I'd like to get a sense of the variability in estimated intercepts and slopes across all my bootstrap samples.

To do that, I'll use the analytic pipeline we established above: fit the model; tidy the output; unnest and examine the results. The code chunk below uses this pipeline to look at bootstrap standard errors for the estimated regression coefficients. 

```{r}
bootstrap_results = 
  boot_straps %>% 
  mutate(models = map(strap_sample, ~lm(y ~ x, data = .x) ),
         results = map(models, broom::tidy)) %>% 
  select(-strap_sample, -models) %>% 
  unnest() %>% 
  group_by(term) %>% 
  summarize(boot_se = sd(estimate))

bootstrap_results
```

Comparing these to the results of ordinary least squares, we see that the standard error for the intercept is much smaller and the standard error for the intercept is a bit larger. This is reasonable, given the non-constant variance in the data given smaller residuals around zero and larger residuals in the the tails of the `x` distribution.

In this case, we can expand the three-panel plot we showed previously to visualize the results of our bootstrap process. 

```{r}
boot_straps %>% 
  unnest() %>% 
  ggplot(aes(x = x, y = y)) + 
  geom_line(aes(group = strap_number), stat = "smooth", method = "lm", se = FALSE, alpha = .1, color = "blue") +
  geom_point(data = sim_df_nonconst, alpha = .5)
```

In comparison to the standard error bands in our previous plot (which are based on OLS), the distribution of regression lines is narrower near $x = 0$ and wider at the ends of the $x$ distribution.

### `bootstrap`

Bootstrapping is common enough that it's been automated, to some degree, in the `modelr::boostrap` function. This function makes it easy to draw bootstrap samples, and stores them in a mostly-helpful way -- as a `resample` object that can be converted to and treated like a data frame.

```{r}
library(modelr)

boot_straps = 
  sim_df_nonconst %>% 
  bootstrap(n = 1000)

boot_straps$strap[[1]]
as_data_frame(boot_straps$strap[[1]])
```

Let's repeat our analysis pipeline using the `bootstrap` function instead of our own process for drawing samples with replacement.

```{r}
sim_df_nonconst %>% 
  bootstrap(n = 1000) %>% 
  mutate(models = map(strap, ~lm(y ~ x, data = .x) ),
         results = map(models, broom::tidy)) %>% 
  select(-strap, -models) %>% 
  unnest() %>% 
  group_by(term) %>% 
  summarize(boot_se = sd(estimate))
```

The results are the same (up to resampling variability), and the code to get here is pretty clean.

Also, check this out -- to bootstrap the dataset with constant error variance, we only have to change the input dataframe!

```{r}
sim_df_const %>% 
  bootstrap(n = 1000) %>% 
  mutate(models = map(strap, ~lm(y ~ x, data = .x) ),
         results = map(models, broom::tidy)) %>% 
  select(-strap, -models) %>% 
  unnest() %>% 
  group_by(term) %>% 
  summarize(boot_se = sd(estimate))
```

These results generally agree with the output of the OLS procedure, which is nice. 

## Other materials

List columns take some getting used to; there are some materials to help with that.

* R for Data Science has a chapter on [fitting many models](http://r4ds.had.co.nz/many-models.html)
* Jenny Bryan's [purrr tutorial](https://jennybc.github.io/purrr-tutorial/) has useful list-column examples

Boostrapping and resampling are also new concepts; the materials below explore these using tidyverse approaches. 

* The [`modelr` package](https://github.com/tidyverse/modelr) has a page
* The bootsrapping [vignette](https://cran.r-project.org/web/packages/broom/vignettes/bootstrapping.html) uses a framework similar to the one we used
* We didn't discuss cross validation, another popular approach, ut you can read up on it [here](https://drsimonj.svbtle.com/k-fold-cross-validation-with-modelr-and-broom) and [here](http://rpubs.com/dgrtwo/cv-modelr)

The code that I produced working examples in lecture is [here](https://github.com/jeff-goldsmith/example_iteration).
