---
title: "Visualization with `ggplot2`"
subtitle: "Part 2"
output:
  html_document: 
    toc: true
    toc_float: true
---

Good visualization is a critical step in data analysis. 

This is the second module in the [Visualization and EDA](topic_visualization_and_eda.html) topic; the relevant slack channel is [here](https://p8105-fall2018.slack.com/messages/CCCBZPUVC). Note that the slides and Other Materials / Extra Reading is the same as in [Visualization Pt 1](visualization_pt1.html).

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

library(tidyverse)
theme_set(theme_gray())
```

## Some slides

<script async class="speakerdeck-embed" data-id="7272d2325e0d4f27a6bca25afbf149ce" data-ratio="1.77777777777778" src="//speakerdeck.com/assets/embed.js"></script>
<div style="margin-bottom:5px"> <strong> <a href="https://speakerdeck.com/jeffgoldsmith/dsi-visualization" title="Visualization" target="_blank">Visualization</a> </strong> from <strong><a href="https://speakerdeck.com/jeffgoldsmith" target="_blank">Jeff Goldsmith</a></strong>. </div><br>


## Example


Although I could add to the same .Rmd I started in [Visualization Pt 1](visualization_pt1.html), I'll create a new document for today. We're also going to take advantage of features in additional packages; most of these were installed previously, but you might need to install some others using the code below:

```{r, eval = FALSE}
devtools::install_github("patchwork")
devtools::install_github("gganimate")
```



```{r}
library(tidyverse)
library(ggridges)
library(ggthemes)
library(patchwork)
# library(gganimate)
```

We'll still work with NOAA weather data, which is loaded using the same code as in [Visualization Pt 1](visualization_pt1.html). 

```{r weather_data_create, cache = TRUE}
library(rnoaa)

weather_df = 
  rnoaa::meteo_pull_monitors(c("USW00094728", "USC00519397", "USS0023B17S"),
                      var = c("PRCP", "TMIN", "TMAX"), 
                      date_min = "2017-01-01",
                      date_max = "2017-12-31") %>%
  mutate(
    name = recode(id, USW00094728 = "CentralPark_NY", 
                      USC00519397 = "Waikiki_HA",
                      USS0023B17S = "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10) %>%
  select(name, id, everything())
weather_df
```

Start with basic scatterplot:

```{r}
ggplot(weather_df, aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5)
```


### Themes and colors

There are a variety of useful ways to change the appearance of your plot, especially if your graphic is intended to be viewed by others. Some of the most important are the axis labels, title, and caption, all of which can be controlled using `labs()`.

```{r}
ggplot(weather_df, aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5) + 
  labs(
    title = "Temperature plot",
    x = "Minimum daily temperature (C)",
    y = "Maxiumum daily temperature (C)",
    caption = "Data from the rnoaa package"
  )
```

Another that I frequently change is the legend position; by default this is on the right of the graphic, but I often shift it to the bottom to ensure the graphic takes up the available left-to-right space. 

```{r}
ggplot(weather_df, aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5) + 
  labs(
    title = "Temperature plot",
    x = "Minimum daily temperature (C)",
    y = "Maxiumum daily temperature (C)",
    caption = "Data from the rnoaa package"
  ) + 
  theme(legend.position = "bottom")
```

While we're on the subject, you can change the default theme (which is `theme_gray`) to something else. Here's `theme_bw()`:

```{r}
ggplot(weather_df, aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5) + 
  labs(
    title = "Temperature plot",
    x = "Minimum daily temperature (C)",
    y = "Maxiumum daily temperature (C)",
    caption = "Data from the rnoaa package"
  ) + 
  theme_bw() + 
  theme(legend.position = "bottom")
```

... and here's `theme_classic()`:

```{r}
ggplot(weather_df, aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5) + 
  labs(
    title = "Temperature plot",
    x = "Minimum daily temperature (C)",
    y = "Maxiumum daily temperature (C)",
    caption = "Data from the rnoaa package"
  ) + 
  theme_classic() + 
  theme(legend.position = "bottom")
```

... and, for some reason, here's the Excel 2003 theme from [`ggthemes`](https://cran.r-project.org/web/packages/ggthemes/vignettes/ggthemes.html):

```{r}
ggplot(weather_df, aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5) + 
  labs(
    title = "Temperature plot",
    x = "Minimum daily temperature (C)",
    y = "Maxiumum daily temperature (C)",
    caption = "Data from the rnoaa package"
  ) + 
  ggthemes::theme_excel() + 
  theme(legend.position = "bottom")
```

Don't use the Excel 2003 theme (the first two are fine, and `ggthemes` has other very nice themes as well). 

The behavior of your plot depends on the data you've supplied; in some cases, it's easier to control behavior through data manipulation than it is through the plot code. This is particularly true for the order of categorical or factor variables. Categorical variables will be ordered alphabetically; factors will be follow the specified order level. You can change the order level of a factor variable to your specified preference using `forcats::fct_relevel` or according to the value of another variable using `forcats::fct_reorder`. 

```{r}
weather_df %>%
  mutate(name = forcats::fct_relevel(name, c("Waikiki_HA", "CentralPark_NY", "Waterhole_WA"))) %>% 
  ggplot(aes(x = name, y = tmax)) + 
  geom_violin(aes(fill = name), color = "blue", alpha = .5) + 
  stat_summary(fun.y = median, geom = "point", color = "blue", size = 4) + 
  theme(legend.position = "bottom")
```

```{r}
weather_df %>%
  mutate(name = forcats::fct_reorder(name, tmax)) %>% 
  ggplot(aes(x = name, y = tmax)) + 
  geom_violin(aes(fill = name), color = "blue", alpha = .5) + 
  stat_summary(fun.y = median, geom = "point", color = "blue", size = 4) + 
  theme(legend.position = "bottom")
```

We'll learn more about the [`forcats`](http://forcats.tidyverse.org/index.html) package in [Data Wrangling II](data_wrangling_ii.html). 

**_Learning Assessment:_** Revisit to density plot you made comparing precipitation across locations. Use themes and labels to improve the readability of this plot. 

### Axis labels


### `geom` data argument

split weather into two

### Tips for polishing figures

Note: i never remember any of this, and just google it repeatedly

Scales

Tick marks

Legend location

"Final" global options I set

Rotating axis text


### `patchwork`

tmax vs tmin + precip ridge


### Examples

PULSE data



## Other materials

Oh man is there a lot of stuff about visualization ...

* There are overviews on good and bad graphics
    * Including an early paper on ["How to display data badly"](http://www.jstor.org/stable/2683253?seq=1#page_scan_tab_contents)
    * Karl Broman's [top ten worst graphs](https://www.biostat.wisc.edu/~kbroman/topten_worstgraphs/)
    * ... and Karl's talk on [creating effective figures and table](https://www.biostat.wisc.edu/~kbroman/presentations/graphs2017.pdf)
    * Also Hadley Wickham's [paper](http://vita.had.co.nz/papers/layered-grammar.pdf) on the philosophy underlying `ggplot`
* There are tutorials on making graphics using `ggplot`
    * From R for Data Science: [basics](http://r4ds.had.co.nz/data-visualisation.html) and [advanced stuff](and http://r4ds.had.co.nz/graphics-for-communication.html)
    * Jenny Bryan's [intro to ggplot](http://stat545.com/cm006_tibbles-dplyr-ggplot2.html) and her [ggplot tutorial](https://github.com/jennybc/ggplot2-tutorial) (with a [video presentation](https://www.youtube.com/watch?v=4MfUCX_KpdE) of the ggplot2 tutorial slides) 
    * From [R Programming for Research](https://geanders.github.io/RProgrammingForResearch/exploring-data-1.html#plots-to-explore-data)
    * The [Graphs](http://www.cookbook-r.com/Graphs/) chapter in the R Cookbook by Winston Chang
    * ... and his [R Graphics Cookbook](http://shop.oreilly.com/product/0636920023135.do)
    * And, of course, a [cheatsheet](https://www.rstudio.com/wp-content/uploads/2016/11/ggplot2-cheatsheet-2.1.pdf)
* There are arguments about ggplot vs base R graphics
    * [Why I don't use ggplot2](https://simplystatistics.org/2016/02/11/why-i-dont-use-ggplot2/)
    * [Comparing ggplot2 and R Base Graphics](https://flowingdata.com/2016/03/22/comparing-ggplot2-and-r-base-graphics/)
    * [Why I use ggplot2](http://varianceexplained.org/r/why-I-use-ggplot2/)


The code that I produced working examples in lecture is [here](./lecture_code/visualization_and_eda.zip).
